Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-23 latex/main.tex
Fix Singer citation

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
of interest in their application in population and statical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
A subtle but important point distinguishing recent work on ARGs
from the classical literature is that the ARG 
``data structure''\cite{Minichiello2006mapping,Wong2024} is no longer
tightly coupled to the generative
process\cite{Hudson1983,Griffiths1997ancestral}, substantially 
broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
Another important point is that strong software support for ARGs now
exists, and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing 
software ecosystem of 
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in 
Figure~\ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants inferred within each week along the ARG,
focusing on the 354 events passing QC filters.
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombinant events rose with the global number of cases
(Figure~\ref{fig:recombinants_over_time}B; 
% JK is this correct?? we had 1.0 + 2.7 \; 10^{-7} x$ which I found confusing
linear regression: $1.0 + 2.7\times 10^{-7} x$, $p = 1.1 \times10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure \ref{fig:recombinants_over_time}C; linear regression: 
$0.9 + 4.8 x$, $p = 3.4\times10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the ``expected heterozygosity'').

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.85\linewidth]{recombinants_over_time.pdf}
\caption{Recombination events across the pandemic. (A) Timing of the
recombination events (black curve, right y-axis) relative to the global case
numbers (\cite{Dong2020}; left y-axis). Cases are color-coded by their Scorpio
labels, with the key showing those groupings with $\geq 1,000$ samples in the
Viridian dataset. Recombinants more commonly occur (B) in weeks with high case
numbers and (C) in weeks with extensive lineage variation, as estimated by high
expected heterozygosity; solid lines show fits by linear regression, allowing
for non-zero intercepts to account for unknown cases and heterozygosity not
captured by the Scorpio labels.}
\label{fig:recombinants_over_time}
\end{figure}

@@

The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
which overwhelmed existing computational approaches,
motivating the rapid development of new methods\cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
the functionality required largely overlaps.
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
and the VCF Zarr data storage format \cite{Czech2025}
to describe relationships among virus samples, building an ARG over time as data becomes available.
By reusing high-quality and high-performance components of these toolkits,
we could develop the sc2ts inference method itself and
perform all the pandemic-scale analyses in this article with Python code alone and
no compiled languages.

To illustrate the power of these libraries, we provide some examples
in which we compare the Usher and sc2ts intersections in tskit format described earlier.
First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
The total aligned dataset for 1,226,587 samples and 27,431 sites shared by the sc2ts and Usher tree sequences
represents 31.34 GB of nucleotide calls.
Of the 80,574,518 missing data calls (Ns) in the alignments,
sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
Of these sc2ts imputed only 77 (0.02\%) incorrectly (i.e., with a base that is not compatible with the ambiguity code).
UShER imputed only 4 calls from this set incorrectly,
reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5 GB
(see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format).
Performing this analysis using matUtils would require exporting the full dataset to VCF,
and converting the VCF to FASTA, and then comparing the alignments one-by-one to the source FASTA alignments.
In contrast with matUtils and most phylogenetic software,
which provide a command line interface and suite of predefined operations that typically output text files,
tskit is a Python API (with C and Rust interfaces also supported) which has many advantages \cite{Baumdicker2022}.
As an illustration of the open-ended and efficient nature of the analyses this enables,
we computed the number of polytomies (nodes that have $> 2$ children)
and the maximum number of children per node in the Usher and sc2ts outputs.
In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
and compute the maximum number of children (5918).
In the sc2ts ARG, it took 385 ms and 6 lines of code to find the maximum number of polytomies (137,001) and
maximum number of children (7708) across all 348 local trees.

@@

During primary inference, gap characters (``-'') are masked 
out as missing data, but this excludes both important information
about deletions from the ARG as well as potentially informative 
phylogenetic signal. 
We therefore performed post-hoc parsinony mapping at 163 sites to include 
the effects of deletions and to ensure that all Pango-informative
sites are included in the ARG.
Using deletions with a 
a minimum frequency of 1\% 
(count of sequences divided by 9,149,680 sequences analyzed)
from Table S1 of Li et al.\cite{Li2023indels},
we chose 75 sites to remap.
In addition, we remapped data at 88 sites that were excluded
from primary inference and are defined as Pango lineage 
defining sites by
 Freyja\cite{Karthikeyan2022}
(\url{https://raw.githubusercontent.com/andersen-lab/Freyja/refs/heads/main/freyja/data/lineage\_mutations.json}).
Alignment data for these 163 sites was then mapped to the 
ARG using the Fitch-Hartigan 
algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
implemented in tskit, including the gap character as a state 
along with the 4 nucleotides.
Following this, we iteratively applied the node parsimony 
heuristics in sc2ts until there were no further reductions in 
the numbers of mutations.

@@

We developed a SnpIt-like~\cite{OToole2024} visualisation 
that shows the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci
(See Fig~\ref{fig:copying_pattern_examples} caption for details).
Inspection of these copying patterns for all the recombination events
identified associations between recombination events that failed QC and
known problematic loci in the SARS-CoV-2 genome.
Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}).
Q1 shows a well supported recombination event that passed QC (albeit marginally),
having a net number of supporting loci of 4 on both sides of the breakpoint.
Q2 shows an artefactual recombination event
caused by a 6-base Delta lineage-defining deletion.
There are 71 (14\%) QC-failing recombination events involving this deletion, and
many cases are affected by a known amplicon dropout
caused by this particular deletion\cite{Sanderson2021}.
Q3 shows an artefactual recombination event caused by
a combination of this Delta deletion and a known MNS (see below).
Finally, Q4 shows an artifactual recombination event
associated with the A507T-T508C-G509A MNS,
which has been identified as a sequencing error
due to incomplete read trimming \cite{DeMaio2024}.
There are 9 recombination events involving this particular MNS, of which 6 failed QC.

@@

We then computed intersection ARGs for sc2ts
(sc2ts\_viridian\_inter\_v1.2.trees.tsz) and UShER 
(usher\_viridian\_inter\_v1.2.trees.tsz) 
by 
simplifying\cite{Kelleher2018efficient,Wong2024} 
both to the 2,475,418 samples and 27,507 sites that are shared
(the UShER tree has mutations at 27,508 sites and sc2ts ARG has 29,893).
As there was no documented means of deriving node dates from 
the UShER tree, and we wish to visually compare 
the phylogenetic backbones, we
also dated the intersection tree by fixing
the sample dates using the ``Date\_tree'' metadata field, 
and using tsdate (version 0.2.3) to date internal nodes.
We did not date the main UShER tree in this way because
of the time-travelling samples present.

@@

\subsubsection*{XS}
https://github.com/jeromekelleher/sc2ts-paper/issues/287
In the ARG, the XS samples appear to have originated from
two sequential recombination events.
However, the second event is likely an artifact,
as the XS sample associated with this event failed
the minimum net number of supporting loci.
In addition, this sample was sequenced using the AmpliSeq V1 kit,
which was also used to sequence eight other samples descending from the same event.
We excluded this likely artifactual event from further analysis and
consider XS to have emerged from a single recombinant origin.

\subsubsection*{XM}
https://github.com/jeromekelleher/sc2ts-paper/issues/284
Possibly related to deletions.

\subsection*{Multiple breakpoints}
Seven of the Pango X lineages present in the Viridian v04 dataset have more than one breakpoint,
with the supporting evidence documented in the associated Pango designation issues.
Five of these seven X lineages were detected as recombinants by sc2ts.
Three of them (XAY, XBC, XBT) had more than one breakpoint,
but were not added to the ARG due to high HMM cost and
a lack of sample support to be incorporated as retrospective sample groups.
However, the other two (XAC and XAK) had only one breakpoint.

\subsubsection*{XAY}
https://github.com/jeromekelleher/sc2ts-paper/issues/490

\subsubsection*{XBC}
https://github.com/jeromekelleher/sc2ts-paper/issues/486
Sc2ts detected XBC as a three-breakpoint recombinant,
correctly identifying the parental Pango lineages.
The inferred breakpoint intervals mostly overlap with
those identified in its designation issue.
XBC was not added to the ARG due to high HMM costs and
inadequate sample support
(within a seven-day retrospective time window).

\subsubsection*{XBT}
https://github.com/jeromekelleher/sc2ts-paper/issues/487
Sc2ts detected XBT as a two-breakpoint recombinant,
correctly identifying the parental Pango lineages.
The inferred breakpoint intervals are identical to those documented in its designation issue.
XBT was not added to the ARG due to high HMM cost and low sample support
(only one QC-passing sample in the Viridian v04 dataset).

\subsubsection*{XAC}
https://github.com/jeromekelleher/sc2ts-paper/issues/357

Evidence of the second breakpoint reported in its designation issue relies on
the presence of a large deletion inherited from BA.2*,
which is currently treated as missing data by sc2ts.
The XAC samples share a recombination node with
the samples of four other Pango X lineages,
which do not have this deletion (Figure 2C).

\subsubsection*{XAK}
https://github.com/jeromekelleher/sc2ts-paper/issues/484
Our results agree with its designation issue in terms of the parental Pango lineages and breakpoint intervals.
XAK was not added to the ARG due to HMM costs (9 and 10) and insufficient sample support (only two samples).
Interestingly, our results suggest that XAK might be a product of two sequential recombination events.

\subsubsection*{Non-recombinants}
\subsubsection*{XN and XAU}

https://github.com/jeromekelleher/sc2ts-paper/issues/348
https://github.com/jeromekelleher/sc2ts-paper/issues/358

Both of these putative recombinants have been defined
on the basis of an RHS from BA.2 and a LHS from BA.1.
In the case of XN, the supposed BA.1 mutations are [C241T, A2832G, C3037T].
In the case of XAU, this set also includes C2470T, so [C241T, C2470T, A2832G, C3037T].
However, both BA.1 and BA.2 are now defined as also having C241T and C3037T.
In addition, a very large number of BA.2 samples also have A2832G, and
three BA.2-designated samples also have C2470T.
This means that there is no longer any evidence for a BA.1 parent for XN, and
only one uncertain mutation that suggests a BA.1 parent for XAU.

\subsubsection*{XAS}
https://github.com/jeromekelleher/sc2ts-paper/issues/340
This was proposed as a recombination between BA.5 (LHS) and BA.2 (RHS).
However, we find that this is only 2 mutations different from
e.g. ERR9929846 which is a BA.4 sample.
In fact, XAS only differs from the BA.4 root by a single mutation, C27945T,
which was identified in the original designation issue
as an important ORF truncation mutation also present in a few BA.2.65 samples.
It is possible that C27945T was acquired by recombination,
this is clearly below our threshold for considering something to be a recombinant.

\subsubsection*{XB}
https://github.com/jeromekelleher/sc2ts-paper/issues/332
XB was proposed to have originated from recombination between B.1.634 (left parent) and B.1.631 (right parent).
None of the samples in Viridian v04 belonging to these Pango lineages were integrated into the ARG due to high HMM costs.

\subsubsection*{XP}
https://github.com/jeromekelleher/sc2ts-paper/issues/345

\subsubsection*{XAJ}
https://github.com/jeromekelleher/sc2ts-paper/issues/352

\subsubsection*{XAZ}
https://github.com/jeromekelleher/sc2ts-paper/issues/356
Proposed as a recombinant of BA.2.5 on the left and BA.5 on the right, with 3 supporting sites (C2232T, C3317T, T3358C).
In sc2ts this is very close to being a recombinant of BA.2.5 (ERR9615610) on the left,
as not only are the 3 mutations present in this sample,
but there is another site (C1912T apparently missed in the designation) which supports recombination.
XAZ has C44T which is lacking in all BA.2.5 samples,
giving only three net supporting loci for a putative recombinant.
However, position 44 is often clipped, so this is not strong evidence against a recombinant, and
this is a very finely balanced case.

\subsubsection*{XAN and XAV}
https://github.com/jeromekelleher/sc2ts-paper/issues/353
Below threshold.

\subsubsection*{XBE}
This was proposed as a recombinant between BA.5,5 and BE.4.
We find two net supporting loci for a recombination breakpoint to be around 22599.

>>>>

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
of interest in their application in population and statical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
A subtle but important point distinguishing recent work on ARGs
from the classical literature is that the ARG 
``data structure''\cite{Minichiello2006mapping,Wong2024} is no longer
tightly coupled to the generative
process\cite{Hudson1983,Griffiths1997ancestral}, substantially 
broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
Another important point is that strong software support for ARGs now
exists, and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing 
software ecosystem of 
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG
(shading in Figure~\ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group
of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants
inferred within each week along the ARG,
focusing on the 354 events passing QC filters.
Recombination events were inferred more often
during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombinant events rose with the global number of cases
(Figure~\ref{fig:recombinants_over_time}B; 
% JK is this correct?? we had 1.0 + 2.7 \; 10^{-7} x$ which I found confusing
linear regression: $1.0 + 2.7\times 10^{-7} x$, $p = 1.1 \times10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure~\ref{fig:recombinants_over_time}C; linear regression: 
$0.9 + 4.8 x$, $p = 3.4\times10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the ``expected heterozygosity'').

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.85\linewidth]{recombinants_over_time}
\caption{
Recombination events across the pandemic.
(A) Timing of the recombination events (black curve, right y-axis)
relative to the global case numbers (\cite{Dong2020}; left y-axis).
Cases are color-coded by their Scorpio labels,
with the key showing those groupings
with $\geq 1,000$ samples in the Viridian dataset.
Recombinants more commonly occur
(B) in weeks with high case numbers and
(C) in weeks with extensive lineage variation,
as estimated by high expected heterozygosity;
solid lines show fits by linear regression,
allowing for non-zero intercepts to account for
unknown cases and heterozygosity not captured by the Scorpio labels.
}
\label{fig:recombinants_over_time}
\end{figure}

@@

The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
which overwhelmed existing computational approaches,
motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
the functionality required largely overlaps.
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
and the VCF Zarr data storage format \cite{Czech2025}
to describe relationships among virus samples,
building an ARG over time as data becomes available.
By reusing high-quality and high-performance components of these toolkits,
we could develop the sc2ts inference method itself and
perform all the pandemic-scale analyses in this article
with Python code alone and no compiled languages.

To illustrate the power of these libraries, we provide some examples
in which we compare the Usher and sc2ts intersections in tskit format described earlier.
First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
The total aligned dataset for 1,226,587 samples and 27,431 sites
shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
Of the 80,574,518 missing data calls (Ns) in the alignments,
sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
Of these sc2ts imputed only 77 (0.02\%) incorrectly
(i.e., with a base that is not compatible with the ambiguity code).
UShER imputed only 4 calls from this set incorrectly,
reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
This computation,
performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
with a peak memory usage of about 4.5 GB
(see \cite{Czech2025} for further benchmarks for
processing the Viridian data in VCF Zarr format).
Performing this analysis using matUtils would require
exporting the full dataset to VCF, and
converting the VCF to FASTA, and
then comparing the alignments one-by-one to the source FASTA alignments.
In contrast with matUtils and most phylogenetic software,
which provide a command line interface and suite of predefined operations that typically output text files,
tskit is a Python API (with C and Rust interfaces also supported)
which has many advantages \cite{Baumdicker2022}.
As an illustration of the open-ended and efficient nature of the analyses this enables,
we computed the number of polytomies (nodes that have $> 2$ children)
and the maximum number of children per node in the Usher and sc2ts outputs.
In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
and compute the maximum number of children (5918).
In the sc2ts ARG, it took 385 ms and 6 lines of code
to find the maximum number of polytomies (137,001) and
maximum number of children (7708) across all 348 local trees.

@@

During primary inference, gap characters (``-'') are masked 
out as missing data, but this excludes both important information
about deletions from the ARG as well as potentially informative 
phylogenetic signal. 
We therefore performed post-hoc parsinony mapping at 163 sites to include 
the effects of deletions and to ensure that all Pango-informative
sites are included in the ARG.
Using deletions with a minimum frequency of 1\%
(count of sequences divided by 9,149,680 sequences analyzed)
from Table S1 of Li et al.\cite{Li2023indels},
we chose 75 sites to remap.
In addition, we remapped data at 88 sites that were excluded
from primary inference and are defined as Pango lineage 
defining sites by Freyja\cite{Karthikeyan2022}
(\url{https://raw.githubusercontent.com/andersen-lab/Freyja/refs/heads/main/freyja/data/lineage\_mutations.json}).
Alignment data for these 163 sites was then mapped to the 
ARG using the Fitch-Hartigan algorithm\cite{Fitch1971toward,Hartigan1973minimum}
implemented in tskit, including the gap character as a state 
along with the four nucleotides.
Following this, we iteratively applied the node parsimony 
heuristics in sc2ts until there were no further reductions in 
the numbers of mutations.

@@

We developed a SnpIt-like~\cite{OToole2024} visualisation 
that shows the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci
(See Figure~\ref{fig:copying_pattern_examples} caption for details).
Inspection of these copying patterns for all the recombination events
identified associations between recombination events that failed QC and
known problematic loci in the SARS-CoV-2 genome.
Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}).
Q1 shows a well supported recombination event that passed QC (albeit marginally),
having a net number of supporting loci of 4 on both sides of the breakpoint.
Q2 shows an artefactual recombination event
caused by a 6-base Delta lineage-defining deletion.
There are 71 (14\%) QC-failing recombination events involving this deletion, and
many cases are affected by a known amplicon dropout
caused by this particular deletion \cite{Sanderson2021}.
Q3 shows an artefactual recombination event caused by
a combination of this Delta deletion and a known MNS (see below).
Finally, Q4 shows an artifactual recombination event
associated with the A507T-T508C-G509A MNS,
which has been identified as a sequencing error
due to incomplete read trimming \cite{DeMaio2024}.
There are 9 recombination events involving this particular MNS,
of which 6 failed QC.

@@

We then computed intersection ARGs for
sc2ts (sc2ts\_viridian\_inter\_v1.2.trees.tsz) and
UShER (usher\_viridian\_inter\_v1.2.trees.tsz)
by simplifying \cite{Kelleher2018efficient,Wong2024}
both to the 2,475,418 samples and 27,507 sites that are shared
(the UShER tree has mutations at 27,508 sites and sc2ts ARG has 29,893).
As there was no documented means of deriving node dates from
the UShER tree, and we wish to visually compare
the phylogenetic backbones,
we also dated the intersection tree by fixing
the sample dates using the ``Date\_tree'' metadata field,
and using tsdate (version 0.2.3) to date internal nodes.
We did not date the main UShER tree in this way because
of the time-travelling samples present.

@@

\subsubsection{XA}
(also illustrated in Figure 2a / 3d)
XA traces to a well supported recombination node, with no reversions.
Note that the “causal” sample ERR5308556
(which triggers the initial recombination event)
is identical to the recombination node, and
therefore lacks C8090T,
which is one of the XA consensus mutations
(identified as those shared by over 90\% of XA samples).
As mentioned elsewhere,
the sc2ts detection results in terms of
the Pango lineages of the parents and the breakpoint intervals
agree with the results reported by Jackson et al. (\ref{tab:jackson}).
Two of the Jackson et al. XA samples
are present in the ARG (ERR5308556 and ERR5414941);
these samples are specifically marked in Figure~\ref{fig:subgraphs},
which is a summary of this subgraph.

\subsubsection{XB}
https://github.com/jeromekelleher/sc2ts-paper/issues/332

The displayed subgraph shows a small selection of the 192 XB samples.
XB was originally defined as a recombinant between
the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
from North and Central America \cite{Gutierrez2022}.
However, these parent lineages were not added to the ARG due to a high HMM cost.
Together with other lineages such as B.1.627
(sister to B.1.631 \cite{Gutierrez2022}),
they are likely to attach somewhere along the long branch
in the subgraph visualization (with >30 mutations),
which also contains the key "R2" deletion
associated with XB / B.1.634 \cite{Gutierrez2022}.

We hypothesise that there are a set of samples
from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
which, should they be added to the ARG, would break up the long branch,
with XB being a recombination between two of them.
In the absence of these, sc2ts cannot detect XB as a recombinant.
Also, note that the XB consensus mutations all occur on or above the long branch.
We suspect that the child node of the long branch
represents the (potentially recombinant) origin of XB, and
the B.1 samples at the bottom left of the subgraph
may be misclassified by pangolin.

Further work is required to determine
why these additional samples were not gradually added
to the ARG by the daily sample matching process or
retrospective sample grouping.

\subsubsection{XC}
The five XC samples descend from a well supported recombination node.
The two parents that are both identical to actual samples
(ERR4908740 on the left and DRR321274 on the right).
There is one reversion (G21987A) on the left side of the breakpoint,
but the site 21987 was masked during ARG inference
because it was identified to be experiencing
frequent mutation in a preliminary ARG.

\subsubsection{XE/XH}
https://github.com/jeromekelleher/sc2ts-paper/issues/337

The XE samples descend from a well supported recombination node,
with one mutation and one reversion (T22792C).
This reversion was mapped onto the ARG as a part of postprocessing, and
could be placed more parsimoniously by a single C22792T that groups together
XH and the lineage leading to the further BA.2 recombination nodes
displayed on the left.
Nevertheless, the exact arrangement of the mutations at position 22792
should not affect the proposed breakpoint,
which is very distant, at position 11283.

*** Should we describe the sequential recombination nodes to the left,
here: note that the lower one is the RE node for XAF. ***

Note that there might be some misalignment immediately
to the right side of the breakpoint
due to a repeated homopolymeric sequence involving 5-base deletions.
Although this problem might obscure the breakpoint position,
there nevertheless is ample evidence for the right parent.
Furthermore, the only two XH samples in the ARG descend
from the same recombination node,
albeit with a number of mutations.

\subsubsection{XF}
XF is a well-supported recombinant between Delta and Omicron BA.1,
with the breakpoint on the far left of the genome.
All XF samples are the unique descendants of
a well supported recombination node,
with both parent nodes (ERR6193241: AY.4 and ERR8042040: BA.1) and
the single child node (ERR8076129) being samples.

\subsubsection{XG}
The three XG samples descend from a reconstructed internal node,
which in turn descends from a well supported recombination node.
The left parent is a sample (SRR20179863), and
the right parent is a reconstructed internal node
assigned to Omicron BA.2,
which is supported by a sample (ERR9124067) that is identical to it.

\subsubsection{XJ}
All the XJ samples descend from a well supported recombination node.
There are 14 samples (including e.g. ERR8683190)
which sc2ts identified as descendants of the same recombination node
but are labeled BA.2 by pangolin (displayed on the left), and
which do not possess the XJ consensus mutations C22792T and C27945T.
These BA.2 samples could be worthy of further investigation.

\subsubsection{XL}
All the XL samples descend from a well supported recombination node,
with three child samples which are identical to it.
The left parent is a sample (ERR7989424), and
the right parent is a reconstructed internal node
assigned to BA.2.5 with a large number of descendant samples.

\subsubsection{XM/XAL}
https://github.com/jeromekelleher/sc2ts-paper/issues/284

XM is a recombinant of BA.1 (left parent) and BA.2 (right parent).
Sc2ts identifies multiple recombinant origins of XM.
One of these origins leads to a dominant clade
(node 1003220 at position 21,595),
which also contains the only three XAL samples.

Independent evidence for multiple XM origins
comes from consideration of a 9-base deletion
in the Spike gene at position 21633,
which is characteristic of BA.2 but not BA.1
(https://github.com/cov-lineages/pango-designation/issues/361).
Most of the XM-labeled samples in ARG have this deletion,
but three (SRR19024311, ERR9447529, and SRR18775874) do not,
suggesting that XM samples trace to at least two recombination events,
one to the right of the deletion
(which leads to the majority of the XM samples), and one to the right.
In the subgraph, it can be seen that one of the samples, SRR18775874,
has been (presumably wrongly) placed
within the main XM deletion-containing clade.
As sc2ts does not use deletions for inference,
deletion remapping (section XXXX) has reverted the deletion in this case,
creating many insertions on the branch above the sample (magenta).
A better use of deletions in sc2ts would likely result
in placing this sample within one of the two XM clades
which do not have the deletion.

\subsubsection{XN/XAU}
(also illustrated in Figure 2c)
https://github.com/jeromekelleher/sc2ts-paper/issues/348
https://github.com/jeromekelleher/sc2ts-paper/issues/358

In the sc2ts ARG, the initial XN sample (ERR8986821)
is well explained as a non-recombinant,
copying from an inferred internal node
assigned to BA.2 with only one mutation.
This internal node is directly supported by the sample ERR9502469.
All the other XN samples descend from ERR8986821.
A small clade containing XAU is the sister to XN, and
there are a series of samples (ERR9453882, ERR9502469, and ERR9823912)
which support the gradual emergence of both groups
from a BA.2 ancestral lineage via accumulation of single mutations.

We believe that the explanation of XN and XAU
in the sc2ts ARG is correct, and
the evidence supporting the original designation of
XN and XAU as recombinants
comes from inaccurate or problematic data sources.
In the case of XN, support for a recombinant BA.1+BA.2 origin
is from three mutations that are claimed to be BA-1-specific:
C241T, A2832G, and C3037T
(https://github.com/cov-lineages/pango-designation/issues/480).
However, the BA.1/BA.2 split
(https://github.com/cov-lineages/pango-designation/issues/361)
does not list C241T and C3037T as BA.1-specific,
meaning that these two mutations are not supportive of recombination.
Moreover, although the remaining mutation
claimed to support a recombinant origin, A2832G,
is typical of BA.1 and not BA.2,
it is seen in several BA.2 samples (e.g. ERR9453882),
supporting the idea that it has independently arisen within BA.2.
Hence, none of the sites used to designate XN as a recombinant reliably do so.

In the case of XAU, a recombinant origin was proposed
on the basis of the same three sites, plus C2470T
(https://github.com/cov-lineages/pango-designation/issues/894).
However, C2470T is not BA.1-specific
(https://github.com/cov-lineages/pango-designation/issues/361).
Moreover, some BA.2-designated samples such as ERR9823912 also have C2470T,
so we do not believe this to be a recombinant-supporting mutation either.

This case illustrates the value of automated ARG reconstruction
for correcting the inevitable errors that arise
when designating recombinants by hand and
using different (and inconsistent) data sources of
Pango lineage-associated mutational constellations.

\subsubsection{XP}
https://github.com/jeromekelleher/sc2ts-paper/issues/345

This is a case where the Pango X lineage is designated
on the basis of two loci on the far right of the genome:
a single nucleotide mutation (A29510C) and
a rare, large deletion at position 29734 in the S2M step loop
(https://github.com/cov-lineages/pango-designation/issues/481),
which is not common enough to map in sc2ts during postprocessing
and hence not shown here.
Sc2ts requires at least four loci on either side
to infer a QC-passing recombinant (section XXXX).
Therefore, unless this deletion were given more weight than a single locus,
sc2ts would not identify XP as a recombinant,
even if deletions were used during ARG inference.
If this deletion were recurrent, the evidence from sc2ts suggests
that XP might not be a recombinant,
because it is only one mutation different from the sample ERR8628084.

\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
https://github.com/jeromekelleher/sc2ts-paper/issues/338

In the ARG, XR, XU, XAA, XAG, and XAM
are nested under the XQ recombinant lineage.
However, upon closer inspection,
the right hand branch above the recombination node appears problematic,
as it displays three reversion mutations,
which revert mutations at positions 22792, 24023, and 25624
in the ancestral lineage just above.
These sites were excluded during ARG inference, and
remapped onto the ARG by parsimony during ARG postprocessing.
We posit that if these three sites were used during ARG inference,
the right parent of the recombination node
would be chosen to be the ancestral BA.2 sample
on the right-hand path with 18 child nodes.
A more parsimonious reconstruction
would then have the XQ clade as a sibling to
a group of (monophyletic) XAA, XAM, XAG, and XR lineages.
This would remove between 4 and 5 reversion mutations.

\subsubsection{XS}
https://github.com/jeromekelleher/sc2ts-paper/issues/287

In the ARG, the XS samples appear to have originated from
two sequential recombination events (separated by a single branch).
However, the later event is likely an artifact,
because the XS sample (SRR18343220) associated with
this event failed detection-level QC
(i.e., the minimum net number of supporting loci;
see the lower copying pattern).
In addition, this sample was sequenced using the Ion AmpliSeq V1 kit,
which was also used to sequence eight other samples
descending from the same event
(clustered due to a shared sequencing error due to AmpliSeq).
These eight samples would likely attach to the nodes
descending from the first recombination event.
We excluded this likely artifactual event from further analysis and
consider XS to have emerged from a single recombinant origin.

\subsubsection{XW}
All the XW samples descend from a well supported recombination node.
Its left parent is a BA.1.1.15-labelled sample (ERR8069658), and
its right parent is a BA.2-labelled sample (ERR9086669).

\subsubsection{XY}
All the XY samples descend from a well supported recombination node,
through a shared reconstructed internal node.
Its left parent is a reconstructed internal node assigned to BA.1.1,
which is supported by a sample (ERR8627048) identical to it.
Its right parent is a BA.2-labeled sample (ERR9399983).
There are a pair of (orange) reversion mutations
within the XY clade at position 23948:
these were not used for inference
but remapped using parsimony during postprocessing, and
indicate that the topology within the XY clade
could potentially be improved
by alternative phylogenetic reconstruction methods.
